An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets

### 2021-10-14 ju-hofmann/ju-hofmann.Rmd
proof read

In this paper we present an open-source implementation of the algorithm of the Congruent Matching Profile Segments (CMPS) method.
@cmps developed the CMPS method for "objective comparison of striated tool marks" and demonstrated its use in some examples of comparing bullet signature correlations.
Although @cmps conceptually described the CMPS algorithm in their paper, the authors did not release an implementation of their method.
Thus, our effort here is to introduce the CMPS method to the `R` [@R] community and provide an open-source, publicly available implementation of the algorithm to use, review, and improve.
Our implementation is made available as part of the `R` package \pkg{cmpsR} on GitHub at \url{https://github.com/willju-wangqian/cmpsR}.

According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "More than 73 percent (73.7) of the homicides for which the FBI received weapons data in 2019 involved the use of firearms" [@fbiucr], and firearm examination could play an important role in all these cases. \hh{XXX the new data for 2020 came out on Sep 27 - homicides are up by close to 30%. We might want to mention that.}
<!-- Firearm examination has  played an important role in  convictions [@fbiucr] in the United States criminal justice system. \hh{Do we have a number or a percentage of cases where firearm examination has played a role? or maybe gun violence? XXX} --> <!-- https://ucr.fbi.gov/crime-in-the-u.s/2019/crime-in-the-u.s.-2019/topic-pages/expanded-homicide --> 
An important task of firearm examination is to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of striation marks acquired during the firing process as bullets are propelled through the barrel.
Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope [@afte].
This practice has been criticized for its lack of objectivity and the associated problem in calculating an error rate [@pcast].

A report published by @nrc states that "[m]uch forensic evidence-including, for example, bite marks and firearm and toolmark identification---is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline." To overcome those criticisms and concerns, researchers have been making an effort to build databases and develop frameworks and algorithms that bring an objective and quantitative assessment into the field [@nistdb, @brundage, @hamby, @Hamby:2019, @song2005, @ChumbleyL_Scott2010VoTM, @aoas, @pmid30444940; @cmps].
Note that this is not an exhaustive list of these efforts.
The notion of bullet signatures is one of the results of these efforts.
Bullet signatures can be extracted from bullet land engraved areas (LEAs), typically one for each land engraved area, and how those bullet signatures are extracted from scans will be discussed in the Background section.
Since bullet signatures can numerically describe the striation marks on the bullet and can be understood by algorithms, they played an important role in bringing objectivity into the field.
In the following sections we will: review the background of bullet signature comparisons, discuss how we followed the idea described by @cmps for the implementation of the CMPS algorithm, propose new metrics based on the CMPS score and a principled evaluation framework for algorithmic results comparison,  and present results of applying our implementation to real data.

@@

The datasets we worked with come from the James Hamby Consecutively Rifled Ruger Barrel Study [@brundage; @hamby; @Hamby:2019], in particular, Hamby set 252 and Hamby set 44.
For each Hamby set, a total of 35 bullets is fired from (the same) ten consecutively manufactured Ruger P-85 pistol barrels.
Two bullets are fired from each barrel, making up a set of 20 reference bullets.
An additional 15 bullets are fired from these ten barrels in a fashion unknown to the study participant.
The aim of the Hamby Study was to have firearms examiners identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves and lands as shown in \autoref{fig:bullet}.
During the firing process, grooves and lands are engraved on a bullet.
Firearms examiners use striation marks on land engraved areas (LEAs) for their visual comparison.
For algorithmic purposes, 3D topographical images of land engraved areas were obtained and stored in x3p format (XML 3-D Surface Profile).
The x3p format provides a standard way of exchanging 2D and 3D profile data.
It conforms to the ISO5436-2 standard [<http://sourceforge.net/p/open-gps/mwiki/X3p/>] adopted by the OpenFMC (Open Forensic Metrology Consortium), a group of firearm forensics researchers who contributes to the establishment of best practices of using metrology in forensic science.
Hamby set 252 was scanned using a NanoFocus lens at 20x magnification with the scan resolution being 1.5625 \textmu m $\times$ 1.5625 \textmu m per pixel.
Hamby set 44 was scanned at the Roy J Carver High-Resolution microscopy lab at Iowa State.
These scans were acquired with a Sensofar Confocal Light Microscope at 20x magnification for a nominal resolution of 0.645 \textmu m $\times$ 0.645 \textmu m per pixel.
Both Hamby set 252 and Hamby set 44 are publicly available on the NIST Ballistics Database Project [@nistdb].

@@

The white horizontal line in \autoref{fig:process}(a) indicates which cross-section was identified by the algorithm to represent the LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the vertical blue lines in \autoref{fig:process}(c) and \autoref{fig:process}(d)).
A non-parametric LOESS smooth [@loess] is fitted to capture the bullet curvature (\autoref{fig:process}(e)) and, finally, the **bullet signature** (\autoref{fig:process}(f)) is obtained as residuals of the cross-section and the fitted smooth.
Note, that in @cmps bullet signatures are referred to as bullet profiles.
However, to avoid confusion, we distinguish the notion of bullet signatures from bullet profiles.
Bullet profiles are shown in panels (b), (c), and (d) of \autoref{fig:process}, while \autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a correct down-stream analysis of the signatures.
To allow for a human-in-the middle inspection and intervention we have provided an interactive web-application, implemented as `R` Shiny App [@shiny], \pkg{bulletinspectR} to identify and correct those errors.
An example of the extraction process with corresponding code and parameter settings can be found on Github at \url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}. Note that the process of extracting signatures might be different from the one used in @cmps because no code or parameter settings are made available publicly.

@@

```{r process, echo=FALSE, out.width=".9\\textwidth", fig.cap="A framework of obtaining a bullet signature. (a) front view of a scanned land engraved area (LEA). The selected crosscut location is indicated by the white line. (b) view of the cross-section of the land engraved area at the white line in (a). (c) the crosscut data plotted in 2D; blue vertical lines indicate the position of left and right grooves. (d) the crosscut data after chopping the left and right grooves. (e) the fitted curvature using LOESS. (f) after removing the curvature from the crosscut data, the bullet signature is obtained"}
knitr::include_graphics("ju-hofmann_files/figure-latex/figure1_v2.png", dpi = 100)
```

@@

Most algorithms for comparing striation marks are based on the digitized signatures and produce a similarity score [@song2005, @ChumbleyL_Scott2010VoTM, @aoas, @pmid30444940].
The congruent matching profile segments (CMPS) algorithm, developed by @cmps for "objective comparison of striated tool marks", is one such algorithm.
The algorithm's main idea is to take a set of consecutive and non-overlapping basis segments from the comparison and for each segment find the "best" registration position on the reference (the other bullet signature) with respect to their cross-correlation values.
From a comparison of these registration positions, a **congruent registration position** is identified, and the number of basis segments taking the congruent registration position is the CMPS score.
Note that researchers in @cmps refers to the origin of basis segments as the reference, but in this paper we refer to it as the comparison.
High CMPS scores are achieved between more similar signatures and are therefore indicative of a same-source pair.
Low scores between pairs of signatures are attributed to different source pairs.
However, a specific threshold of the CMPS score to distinguish between same-source and different-source comparisons is not provided in @cmps, instead the threshold depends on the underlying structure of the data and the choice of parameters. \hh{XXX  In a legal setting this variability is problematic because it allows for situations in which experts could choose parameters based on a wished-for outcome. }
Further research is needed to understand how to determine optimal threshold settings.
The CMPS algorithm can assist firearm examiners with drawing a conclusion about the source of a comparison pair.
Unfortunately, @cmps did not release any code or specific parameter settings for their implementation of the CMPS algorithm.
Thus, in this paper we present an open-source implementation of the CMPS algorithm in the R package \pkg{cmpsR} available from Github \url{https://github.com/willju-wangqian/cmpsR}. This publicly available implementation calculates the CMPS score of a comparison using the following code:

@@

The section "Implementation" will go through the algorithm and show how to use the \pkg{cmpsR} package. 
A further example that illustrates the main points is also included.
The section "Results" presents the results of evaluating the \pkg{cmpsR} package using Hamby set 252 and Hamby set 44. The results from Hamby set 252 are used to verify that our implementation is, at least qualitatively, comparable to the algorithm described in @cmps. Results from Hamby 44 show the need for a further investigation of the parameter choices even in the case of bullets fired from the same barrels. \wj{We suggest new CMPS metrics that summarize land-level CMPS scores and a sum of squares ratio that can be used to evaluate algorithmic results.}
<!-- \hh{We suggest a sum of square ratio ...} -->
The last section covers some final discussion and conclusions.

@@

```{r tiles, echo=FALSE, fig.cap="CMPS scores of all 36 pairwise bullet signature comparisons for two bullets. Land engraving pairs generated by the same land (KM comparisons) are highlighted", out.width=".7\\textwidth", fig.align="center"}
dframe <- dframe %>% mutate(
  landA = paste0("L", landA),
  landB = paste0("L", landB),
  landB = factor(landB, levels = paste0("L", c(2:6,1))),
  bulletA = paste0("Bullet ", bulletA),
  bulletB = paste0("Bullet ", bulletB)
)

@@

As presented in the work of @cmps, researchers applied the CMPS method to scans of one of the Hamby sets.
While it is not explicitly stated in the paper, we presume this to be Hamby 252, as only those scans were publicly available at the time.
In order to show that our implementation of the CMPS algorithm is able to reproduce the results in @cmps and be used for other data sets, we applied our implementation to both Hamby set 252 and Hamby set 44.
For the purpose of reproducibility, we present how we obtained bullet signatures from the Hamby set data and include code in the appendix: for both Hamby 252 and Hamby 44, we started with scans in the form of x3p files in the database.
Following the framework proposed by @aoas, we used the same set of parameters, removed damaged bullet scans, obtained bullet signatures for each bullet land engraving, and removed outliers in bullet signatures.
Note that researchers of @cmps applied the CMPS algorithm to bullet signatures as well but used a framework different from ours.
However, since their work is not open-source, we were not able to follow their framework and were only able to reproduce the results for Hamby set 252 qualitatively.

@@

\hh{XXX Will, the conclusion needs to be a bit more than just a summary of what we did in the paper - think of it as a summary plus either the limitations of our approach, or next steps that should be done but goes beyond this paper.}

>>>>

In this paper we present an open-source implementation of the algorithm of the Congruent Matching Profile Segments (CMPS) method.
@cmps developed the CMPS method for "objective comparison of striated tool marks" and demonstrated its use in some examples of comparing bullet signature correlations.
Although @cmps conceptually describe the CMPS algorithm in their paper, the authors did not release an implementation of their method.
Thus, our effort here is to introduce the CMPS method to the `R` [@R] community and provide an open-source, publicly available implementation of the algorithm to use, review, and improve.
Our implementation is made available as part of the `R` package \pkg{cmpsR} on GitHub at \url{https://github.com/willju-wangqian/cmpsR}.

According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "More than 73 percent (73.7) of the homicides for which the FBI received weapons data in 2019 involved the use of firearms" [@fbiucr], and firearm examination could play an important role in all these cases. \hh{XXX the new data for 2020 came out on Sep 27 - homicides are up by close to 30%. We might want to mention that.}
<!-- Firearm examination has  played an important role in  convictions [@fbiucr] in the United States criminal justice system. \hh{Do we have a number or a percentage of cases where firearm examination has played a role? or maybe gun violence? XXX} --> <!-- https://ucr.fbi.gov/crime-in-the-u.s/2019/crime-in-the-u.s.-2019/topic-pages/expanded-homicide --> 
An important task of firearm examination is to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of striation marks acquired during the firing process as bullets are propelled through the barrel.
Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope [@afte].
This practice has been criticized for its lack of objectivity and the associated problem of determining valid error rates [@pcast].

A report published by the @nrc states that "[m]uch forensic evidence-including, for example, bite marks and firearm and toolmark identification---is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline." To overcome those criticisms and concerns, researchers have been making an effort to build databases and develop frameworks and algorithms that bring an objective and quantitative assessment into the field \hh{XXX could you summarize the main contribution of each of these papers in at most a sentence?} [@nistdb, @brundage, @hamby, @Hamby:2019, @song2005, @ChumbleyL_Scott2010VoTM, @aoas, @pmid30444940; @cmps].
Note that this is not an exhaustive list of these efforts.
The notion of bullet signatures is one of the results of these efforts.
Bullet signatures can be extracted from bullet land engraved areas (LEAs), typically one for each land engraved area, and how those bullet signatures are extracted from scans will be discussed in the Background section.
Since bullet signatures can numerically describe the striation marks on the bullet and can be understood by algorithms, they played an important role in bringing objectivity into the field.
In the following sections we will: review the background of bullet signature comparisons, discuss how we followed the idea described by @cmps for the implementation of the CMPS algorithm, propose new metrics based on the CMPS score and a principled evaluation framework for algorithmic results comparison,  and present results of applying our implementation to real data.

@@

The datasets we worked with come from the James Hamby Consecutively Rifled Ruger Barrel Study [@brundage; @hamby; @Hamby:2019], in particular, Hamby set 252 and Hamby set 44.
For each Hamby set, a total of 35 bullets is fired from (the same) ten consecutively manufactured Ruger P-85 pistol barrels.
Two bullets are fired from each barrel, making up a set of 20 reference bullets.
An additional 15 bullets are fired from these ten barrels in a fashion unknown to the study participant.
The aim of the Hamby Study was to have firearms examiners identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves and lands as shown in \autoref{fig:bullet}.
During the firing process, grooves and lands are engraved on a bullet.
Firearms examiners use striation marks on land engraved areas (LEAs) for their visual comparison.
For algorithmic purposes, 3D topographical images of land engraved areas were obtained and stored in x3p format (XML 3-D Surface Profile).
The x3p format provides a standard way of exchanging 2D and 3D profile data.
It conforms to the ISO5436-2 standard [<http://sourceforge.net/p/open-gps/mwiki/X3p/>] adopted by the OpenFMC (Open Forensic Metrology Consortium), a group of firearm forensics researchers who contributes to the establishment of best practices of using metrology in forensic science.
Hamby set 252 was scanned using a NanoFocus lens at 20x magnification with the scan resolution being 1.5625 \textmu m $\times$ 1.5625 \textmu m per pixel.
Hamby set 44 was scanned at the Roy J Carver High-Resolution microscopy lab at Iowa State.
These scans were acquired with a Sensofar Confocal Light Microscope at 20x magnification for a nominal resolution of 0.645 \textmu m $\times$ 0.645 \textmu m per pixel.
Both Hamby set 252 and Hamby set 44 are publicly available from the NIST Ballistics Database Project [@nistdb].

@@

The thin white horizontal line in \autoref{fig:process}(a) indicates which cross-section was identified by the algorithm to represent the LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the vertical blue lines in \autoref{fig:process}(c) and \autoref{fig:process}(d)).
A non-parametric LOESS smooth [@loess] is fitted to capture the bullet curvature (\autoref{fig:process}(e)) and, finally, the **bullet signature** (\autoref{fig:process}(f)) is obtained as residuals of the cross-section and the fitted smooth.
Note, that in @cmps bullet signatures are referred to as bullet profiles.
However, to avoid confusion, we distinguish the notion of bullet signatures from bullet profiles.
Bullet profiles are shown in panels (b), (c), and (d) of \autoref{fig:process}, while \autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a correct down-stream analysis of the signatures.
To allow for a human-in-the middle inspection and intervention we have provided an interactive web-application, implemented as `R` Shiny App [@shiny], \pkg{bulletinspectR} to identify and correct those errors.
An example of the extraction process with corresponding code and parameter settings can be found on Github at \url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}. Note that the process of extracting signatures might be different from the one used in @cmps because no code or parameter settings are made available publicly.

@@

```{r process, echo=FALSE, out.width=".9\\textwidth", fig.cap="A framework of obtaining a bullet signature. (a) rendering from the 3d topographic scan of a land engraved area (LEA). The selected crosscut location is indicated by a thin white horizontal line. (b) view of the cross-section of the land engraved area at the white line in (a). (c) the crosscut data plotted in 2D; blue vertical lines indicate the position of left and right grooves. (d) the crosscut data after chopping the left and right grooves. (e) the fitted curvature using LOESS. (f) after removing the curvature from the crosscut data, the bullet signature is obtained"}
knitr::include_graphics("ju-hofmann_files/figure-latex/figure1_v2.png", dpi = 100)
```

@@

Most algorithms for comparing striation marks are based on the digitized signatures and produce a similarity score [@song2005, @ChumbleyL_Scott2010VoTM, @aoas, @pmid30444940].
The congruent matching profile segments (CMPS) algorithm, developed by @cmps for "objective comparison of striated tool marks", is one such algorithm.
The algorithm's main idea is to take a set of consecutive and non-overlapping basis segments from the comparison and for each segment find the "best" registration position on the reference (the other bullet signature) with respect to their cross-correlation values.
From a comparison of these registration positions, a **congruent registration position** is identified, and the number of basis segments taking the congruent registration position is the CMPS score.
Note that researchers in @cmps refer to the origin of basis segments as the reference, but in this paper we refer to it as the comparison signature.
High CMPS scores are achieved between more similar signatures and are therefore indicative of a same-source pair.
Low scores between pairs of signatures are attributed to different source pairs.
However, a specific threshold of the CMPS score to distinguish between same-source and different-source comparisons is not provided in @cmps, instead the threshold depends on the underlying structure of the data and the choice of parameters.   In a legal setting this variability is problematic because it allows for situations in which experts could choose parameters based on whether they are witnesses for the defense or the prosecution. 
Further research is needed to understand how to determine optimal threshold settings.
The CMPS algorithm can assist firearm examiners with drawing a conclusion about the source of a comparison pair.
<!--Unfortunately, @cmps did not release any code or specific parameter settings for their implementation of the CMPS algorithm.-->
Thus, in this paper we present an open-source implementation of the CMPS algorithm in the R package \pkg{cmpsR} available from Github \url{https://github.com/willju-wangqian/cmpsR}. This publicly available implementation calculates the CMPS score of a comparison using the following code:

@@

The section "Implementation" will go through the algorithm and show how to use the \pkg{cmpsR} package. 
A further example that illustrates the main points is also included.
In the section on 'Evaluation metrics" we introduce new CMPS metrics that summarize land-level CMPS scores and a sum of squares ratio that can be used to evaluate algorithmic results.
The section "Results" presents the results of evaluating the \pkg{cmpsR} package using Hamby set 252 and Hamby set 44. The results from Hamby set 252 are used to verify that our implementation is, at least qualitatively, comparable to the algorithm described in @cmps. Results from Hamby 44 show the need for a further investigation of the parameter choices even in the case of bullets fired from the same barrels. 
<!-- \hh{We suggest a sum of square ratio ...} -->
The last section covers some final discussion and conclusions.

@@

```{r tiles, echo=FALSE, fig.cap="CMPS scores of all 36 pairwise bullet signature comparisons for two bullets. Land engraving pairs generated by the same land (KM comparisons) are highlighted. \\hh{Note that in this example the axis along Bullet 2 starts with Land 2. This corresponds to Phase 2 in equation XXX}", out.width=".7\\textwidth", fig.align="center"}
dframe <- dframe %>% mutate(
  landA = paste0("L", landA),
  landB = paste0("L", landB),
  landB = factor(landB, levels = paste0("L", c(2:6,1))),
  bulletA = paste0("Bullet ", bulletA),
  bulletB = paste0("Bullet ", bulletB)
)

@@

As presented in the work of @cmps, researchers applied the CMPS method to scans of one of the Hamby sets.
While it is not explicitly stated in the paper, we presume this to be Hamby 252, as only those scans were publicly available at the time. \hh{XXX Let's take out this statement once we hear back from NIST}
In order to show that our implementation of the CMPS algorithm is able to reproduce the results in @cmps and be used for other data sets, we applied our implementation to both Hamby set 252 and Hamby set 44.
For the purpose of reproducibility, we present how we obtained bullet signatures from the Hamby set data and include code in the appendix: for both Hamby 252 and Hamby 44, we started with scans in the form of x3p files in the database.
Following the framework proposed by @aoas, we used the same set of parameters, removed damaged bullet scans, obtained bullet signatures for each bullet land engraving, and removed outliers in bullet signatures.
Note that researchers of @cmps applied the CMPS algorithm to bullet signatures as well but used a framework different from ours.
However, since their work is not open-source, we were not able to follow their framework and were only able to reproduce the results for Hamby set 252 qualitatively.

@@

\hh{XXX Additional line for the conclusion: Hamby sets 44 and 252 were taken with different resolution scanning devices. The bullets for Hamby 44 and Hamby 252 are fired from the same ten consecutively rifled P-85 Ruger barrels. The difference in parameter settings is therefore quite surprising. As next steps we need validation studies - trying out the CMPS algorithm on firearms and ammunition combinations that are different to test the limits of the algorithm. XXX shoutout to Susan's paper? - will look up the reference }
